Multisequence algorithm for coarse-grained biomolecular simulations: application to protein fold switching

### 2017-05-19 Manuscript.tex
Modified abstract and discussion

\title[Multisequence Monte Carlo simulations]{Multisequence algorithm for coarse-grained biomolecular simulations: application to protein folding switching}

@@

\begin{abstract}
We evaluate a generalized-ensemble algorithm for coarse-grained simulations of biomolecules that allows the thermodynamics of multiple sequences to be calculated in a single ``multisequence" run. By carrying out a random walk in sequence space, conformational sampling is also enhanced. Escape from local energy minima is achieved by visiting sequences for which the minima is shallower or absent. We apply the algorithm to a coarse-grained model for protein folding with 7 atoms per amino acid and apply it to sets of sequences with up to 35 amino acids spanning sequence space  With the multisequence algorithm, we show that the folding thermodynamics of $>$1,000 sequences at relatively low temperatures can be readily calculated. We use the resulting data to explore the statistics of pathways in sequence space between the two pairs of folds.

@@

\keywords{Generalized-ensemble algorithm, simulated tempering, protein folding, protein fold switching, coarse-grained model}

@@

\section{Discussion}
\noindent
We have tested an algorithm for coarse-grained biomolecular simulations that relies on making the sequence of the molecule a dynamical parameter.

For comparison, we used the well-established generalized-ensemble ST algorithm and focused on 144 model sequences with 16 amino acids ({\SR}). As a measure of computational efficiency, we used on the statistical error $\sigE$ in the estimation of the average total energy $E$ for 4 different sequences which are part of {\SR}.

Our results show that at high $T$, well above the folding temperatures of both A1 and N1, we find that the statistical errors $\sigE$ for ST and MS are small and very similar. This is not surprising given that conformational sampling of relatively short polymer chains at high $T$ does not involve crossing major energy barriers. Therefore, successive sampled conformational states, for a given combination of sequence $\sbar$ and temperature $T$, are likely uncorrelated in both ST and MS simulations which, in turn, leads to similar estimates of $\Eave$ (and $\sigE$). At low $T$,...

Hence, we conclude that the computational efficiency of the MS algorithm is comparable to ST. However, that while the ST simulation requires that the thermodynamic behavior is calculated at a range of temperatures, MS works well on its own directly at low temperatures. ST was introduced as an extension to the standard, fixed-$T$ Metropolis-Hastings algorithm in order to overcome the poor mixing of states at low temperatures, i.e., the tendency for the simulations at low $T$ to become trapped in local free energy minima. The MS algorithm, however, efficiently calculates the thermodynamics at low $T$'s without visits to other $T$'s. In the MS algorithm, it is visits to sequences with poor native state stability rather than high $T$'s that drive the decorrelation of generated states and escape from local traps. This way, calculation of low-$T$ thermodynamics is enhanced in MS simulations while at the same time enhancing coverage in sequence space. This makes the MS algorithm ideal for situations in which the finite-$T$ thermodynamic behavior of multiple sequences are of interest.

A notable feature of the MS algorithm is that provides escape from local energy minima on the free energy landscape of proteins without resorting to a change in temperature. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in lipid membranes, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Sampling in membrane simulations at high $T$, such as in replica-exchange simulations, are usually avoided in membrane simulations because it can lead to unwanted perturbation of the lipid bilayer structure although it can be a viable approach in particular cases.~\cite{Ulmschneider2010}

>>>>

\title[Multisequence Monte Carlo simulations]{Multisequence algorithm for coarse-grained biomolecular simulations: application to protein fold switching}

@@

\begin{abstract}
We consider a generalized-ensemble algorithm for coarse-grained simulations of biomolecules, which allows the thermodynamic behaviors of two or more sequences to be determined in a single multisequence run. By carrying out a random walk in sequence space, the method also enhances conformational sampling. Escape from local energy minima is accelerated by visiting sequences for which the minima are more shallow or absent. We apply the method to an intermediate-resolution coarse-grained model for protein folding with 7 atoms per amino acids. The potential for large-scale coverage of sequence space is explored by applying the method to two sets with $>$1,000 sequences each. The resulting thermodynamic data is used to explore the sequence-structure and sequence-stability relationships between pairs of protein folds with different secondary structures. The potential implications for the evolution of new protein folds are discussed.

@@

\keywords{Generalized ensembles, simulated tempering, protein folding, coarse-graining, evolution}

@@

\section{Discussion}
\noindent
We have evaluated a generalized-ensemble algorithm for coarse-grained simulations of biomolecules that works by making the sequence a dynamic parameter and thereby leading to a random walk in sequence space. We find that this approach has two main benefits. On the one hand, it provides a convenient way to sample the respective canonical distributions for a set of sequences in a single run and, on the other, it also enhances the sampling of conformational space.

The sampling efficiency of the MS algorithm can be assessed by the comparison with ST in Figure~3. At the highest studied temperatures, which are well above the folding temperatures of both A1 and N1, the statistical errors on the average total energy, $\sigE$, are very similar for the two methods. This is not unexpected because conformational sampling of short polymers at high $T$ does not involve crossing any major energy barriers. As a result, successively sampled conformations, for a given combination of sequence $\sbar$ and temperature $T$, are therefore likely mostly uncorrelated in both methods which in turn leads to similar $\sigE$ values.

At lower temperatures, we find that the MS simulations often yields significantly smaller $\sigE$ than ST simulations. It is notable that this acceleration in conformational sampling is achieved despite that simulations are carried out at constant temperature. Hence, rather than promoting escape from local minima in the free energy landscape by visits to higher temperature, as in ST~\cite{Marinari1992,Lyubartsev1992} or replica-exchange,~\cite{Swendsen1986} the MS algorithm promotes local minima escape through visits to other sequences for which the minima may not be as severe or are absent altogether. For the same reason, the efficiency of the MS algorithm likely depends on the size and character of sequences included in the simulation. In particular, the MS simulation is likely dependent on the inclusion of at least some less stable sequences to promote at least partial unfolding of the chain and the transition to different conformational states.

Clearly, the MS method should not be seen as a general simulation method to speed up conformational sampling due to the complexity of performing sequence updates in most biomolecular models. However, our results indicate that for coarse-grained models that permit the sequence updates as a Metropolis step and when visits to higher $T$ is unwanted (or unnecessary), the MS method can be highly efficient. A notable feature of the algorithm is that it provides escape from local minima on the free energy landscape of proteins without resorting to a change in $T$. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in the context or ordered dense phases such as lipid bilayers, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Simulations of lipid bilayers at elevated $T$ must typically be avoided because it can lead to unwanted perturbations of the membrane structure, although it can be a viable approach to speed up conformational sampling in particular cases.~\cite{Ulmschneider2010}

%Overall our results show that, when technically feasible, a multisequence approach can be highly efficient for calculating the thermodynamics of a large number of amino acid sequences even at low temperatures. To demonstrate that the method to large sequence sets we applied it to the full binary sequences spaces between the model sequence pairs A1-N1 and A2-TN, which both include 1,024 sequences. These calculations would have been difficult to carry out ``in series" using some other standard method such as simulated tempering.

%Hence, we conclude that the computational efficiency of the MS algorithm is comparable to ST. However, that while the ST simulation requires that the thermodynamic behavior is calculated at a range of temperatures, MS works well on its own directly at low temperatures. ST was introduced as an extension to the standard, fixed-$T$ Metropolis-Hastings algorithm in order to overcome the poor mixing of states at low temperatures, i.e., the tendency for the simulations at low $T$ to become trapped in local free energy minima. The MS algorithm, however, efficiently calculates the thermodynamics at low $T$'s without visits to other $T$'s. In the MS algorithm, it is visits to sequences with poor native state stability rather than high $T$'s that drive the decorrelation of generated states and escape from local traps. This way, calculation of low-$T$ thermodynamics is enhanced in MS simulations while at the same time enhancing coverage in sequence space. This makes the MS algorithm ideal for situations in which the finite-$T$ thermodynamic behavior of multiple sequences are of interest.

%A notable feature of the MS algorithm is that provides escape from local energy minima on the free energy landscape of proteins without resorting to a change in temperature. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in lipid membranes, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Sampling in membrane simulations at high $T$, such as in replica-exchange simulations, are usually avoided in membrane simulations because it can lead to unwanted perturbation of the lipid bilayer structure although it can be a viable approach in particular cases.~\cite{Ulmschneider2010}
